% ===Lesson 12: Citations and references
\documentclass{article}
\usepackage[T1]{fontenc}
\usepackage[margin=1in]{geometry}
\usepackage{verbatim}
\usepackage[parfill]{parskip}
\usepackage{mdframed}

\begin{document}
{\bfseries This lesson show the basics of reference databases. Learn how to build your own databases and how to use them in documents using the two major workflows available.}

\section{Reference databases}
Reference databases are normally referred to as ‘BibTeX files’ and have the extension \verb|.bib|. They contain one or more entries, one for each reference, and within each entry there are a series of fields. Let us look at an example.
\vspace{0.5cm}

\begin{mdframed}
\begin{verbatim}
@article{Thomas2008,
  author  = {Thomas, Christine M. and Liu, Tianbiao and Hall, Michael B.
             and Darensbourg, Marcetta Y.},
  title   = {Series of Mixed Valent {Fe(II)Fe(I)} Complexes That Model the
             {H(OX)} State of [{FeFe}]Hydrogenase: Redox Properties,
             Density-Functional Theory Investigation, and Reactivity with
             Extrinsic {CO}},
  journal = {Inorg. Chem.},
  year    = {2008},
  volume  = {47},
  number  = {15},
  pages   = {7009-7024},
  doi     = {10.1021/ic800654a},
}
@book{Graham1995,
  author    = {Ronald L. Graham and Donald E. Knuth and Oren Patashnik},
  title     = {Concrete Mathematics},
  publisher = {Addison-Wesley},
  year      = {1995},
}
\end{verbatim}
\end{mdframed}
\vspace{0.5cm}

This is an entry for an article and another for a book; these are by far the most common types. Each database entry type starts with \verb|@|, as shown, and all of the information then sits within a brace pair.

Exactly which fields you need to give depends on the type of entry, but most of these are quite obvious. You might notice that in the \verb|author| field, each entry is separated by \verb|and|. This is essential: the format of the output needs to know which author is which. You might also notice that in the article title, some entries are in an extra set of braces; these are there to prevent any case-changing being applied.

\section{Transferring information from the database}

\end{document}